---
title: "dplyr Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
# library(tidytuesdayR)

knitr::opts_chunk$set(echo = FALSE)
```


## Welcome!

### Tidy Tuesday

[Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) is a community-led
effort to practice quick iterations of processing data in R to create data
visualizations and other cool things. There's even a package to make obtaining
the data each week super easy.

```{r}
# tuesdata <- tt_load(2020, week = 9)

# measles <- tuesdata$measles

mtcars
```

## `dplyr`'s 5 Verbs

### `select`

### `filter`



## Data Exploration

```{r}
mtcars
```

### Counts

```{r}
count(mtcars)

count(mtcars) - n_distinct(mtcars)
```


### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

